\vspace{-.5cm} Lights: a generalized joint model for high-dimensional multivariate longitudinal data and censored durations \vspace{.5cm}

### 2022-01-31 manuscript/paper.tex
Update Monte-Carlo stuff

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is from both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
for time-independent feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model : this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model, and yields sparsity for a given trajectory.
Inference is achieved using an efficient and novel proximal Quasi-Newton Monte Carlo Expectation Maximization algorithm (prox-QNMCEM).
Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters.
Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for real-time decision support in a customer's satisfaction monitoring context for instance, or else for clinical decision making in patient monitoring.

@@

\subsection{A Quasi-Newton Monte Carlo EM}
\label{sec:prox-QNMCEM}

@@

\paragraph*{Monte Carlo E-step.}

@@

\paragraph*{Proximal Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNMCEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big( \hat \pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:prox-QNMCEM}) with every increase of $N$.

@@

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \hat \E_{\theta^{(w)}}[ \gamma_k^\top \Psi_{k,i}(t_i) | t_i, \delta_i, y_i] \\
  & - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[\exp \big\{\gamma_k^\top \Psi_{k,i}(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Big[ \delta_i  \hat \E_{\theta^{(w)}}[ \varphi_{k,a}(t_i, b_i^l) | t_i, \delta_i, y_i] \\ \nonumber
   &\qquad\qquad - \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \hat \E_{\theta^{(w)}}[ \varphi_{k,a}(t_j^u, b_i^l) \exp \big\{\gamma_k^\top \Psi_{k,i}(t_j^u)\big\} | t_i, \delta_i, y_i] \ind{\{t_j^u \leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at each uncensored observed event times as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \hat \E_{\theta^{(w)}}\Big[\exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_{k,i}(t)\big\} | t_i, \delta_i, y_i \Big] \ind{\{t_i \geq t\}} },
\end{equation}
for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

Finally, for the update of $\phi^{(w)}$ and regarding Table~\ref{table:glm}, we focus on the Gaussian case being particularly useful in practice. It is obtained in closed-form by
\begin{align}
  \label{eq:update_phi}
  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} &\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ \big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
\end{align}
Appendix~\ref{sec:prox-QNMCEM class} gives some details about the way we implement the equations and their computation in practice.

\paragraph*{Initialization.}
Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced. We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:prox-QNMCEM}. 
For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_k}^{(0)} = \mathbf{0}_{L\sA}$. Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}. Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (details on the algorithm are given in Appendix~\ref{sec:MLMM}, and details on the implementation in Appendix~\ref{sec:MLMM class}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.

\paragraph*{The prox-QNMCEM algorithm.}
Algorithm~\ref{alg:prox-QNMCEM} describes the main steps of the resulting QNMCEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.

@@

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{prox-QNMCEM algorithm for the lights model inference}
\label{alg:prox-QNMCEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{Monte Carlo E-step}
  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:hat_Lambda_ik} for the required functions $\tilde g$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:update_beta} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}
Let us define the relative difference at step $(w+1)$ as
\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
Then in practice, the convergence for Algorithm~\ref{alg:prox-QNMCEM} is declared when the condition
\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
is satisfied for 3 consecutive iterations, since the monotonicity property is lost due to the Monte Carlo approximations.

@@

The question now is to choose a discriminative marker rule for risk prediction to be used in the cross-validation procedure for selecting the best regularization hyper-parameters, and of course to be used as final risk prediction after hyper-parameters fine-tuning.
The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNMCEM algorithm introduced in Section~\ref{sec:prox-QNMCEM}), posterior risk classification for subject $i$ and group $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}). Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.

@@

In this paper, a generalized joint model for high-dimensional multivariate longitudinal data and censored durations (lights) has been introduced, and a new efficient estimation
algorithm (prox-QNMCEM) has been derived, that considers a penalization of the likelihood in order to perform feature
selection and to prevent overfitting.

>>>>

In this paper, we propose a method called \textit{lights} (generaLized joInt hiGH-dimensional longiTudinal Survival) which is from both JLCMs and SREMs, since we also include features extracted from the longitudinal processes as potential risk factor in the survival model, which is a group-specific Cox model~\citep{Cox1972JRSS} with high-dimensional shared associations.
To allow flexibility in modeling the dependency between the longitudinal features and the event time, we use appropriate penalties : elastic net~\citep{zou2005regularization}
for time-independent feature selection in the latent class membership, and sparse group lasso~\citep{simon2013sparse} in the survival model : this penalty acts like the lasso at the trajectory level, namely an entire trajectory may drop out of the model, and yields sparsity for a given trajectory.
Inference is achieved using an efficient and novel proximal Quasi-Newton Expectation Maximization algorithm (prox-QNEM).
Posterior estimates of the latent class membership probabilities given the longitudinal process and the fact that the subject is still alive at prediction time -- in a so called ``real-time prediction paradigm'' -- are then used as discriminative marker rule for risk prediction in the cross-validation procedure for selecting the best regularization hyper-parameters.
Hence, the method provides interpretations of the high-dimensional longitudinal features, thus offering a powerful tool for real-time decision support in a customer's satisfaction monitoring context for instance, or else for clinical decision making in patient monitoring.

@@

\subsection{A Quasi-Newton EM}
\label{sec:prox-QNEM}

@@

\paragraph*{E-step.}

@@

\paragraph*{Proximal Quasi-Newton M-step.}
Here, we need to compute \[\theta^{(w+1)} \in \argmin_{\theta \in \R^\vartheta} \cQ_n(\theta, \theta^{(w)}) + \sum_{k=0}^{K-1} \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta} + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.\]
We then present the parameters updates in the order given in Algorithm~\ref{alg:prox-QNEM}.
First, the update of $D^{(w)}$ is naturally given in closed-form by
\begin{equation}
  \label{eq:update_D}
  D^{(w+1)} = n^{-1} \sum_{i=1}^n \hat \E_{\theta^{(w)}}[ b_i b_i^\top | t_i, \delta_i, y_i].
\end{equation}
Then, let us focus on the update of $\xi_k^{(w)}$ for $k=0, \ldots, K-1$. We denote
\[ P^{(w)}_{n, k}(\xi_k) = -n^{-1} \sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \log \pi_{\xi_k}(x_i) \]
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\xi_k$. The update for $\xi_k^{(w)}$ therefore requires to solve the following convex minimization problem
\begin{equation}
  \label{eq:minimization-xi_k}
  \xi_k^{(w+1)} \in \argmin_{\xi_k \in \R^p} P^{(w)}_{n,k}(\xi_k) + \zeta_{1,k} \norm{\xi_k}_{\text{en}, \eta}.
\end{equation}
It looks like the logistic regression objective, where labels are not fixed but softly encoded by the expectation step (computation of $\hat \pi_{ik}^{\theta^{(w)}}$ in~\eqref{eq:hat_pi_ik-def}). We then choose to solve~\eqref{eq:minimization-xi_k} using the L-BFGS-B algorithm~\citep{zhu1997algorithm} which belongs to the class of quasi-Newton optimization routines and solves the given minimization problem by computing approximations
of the inverse Hessian matrix of the objective function. It can deal with differentiable convex objectives with box constraints.
In order to use it with $\ell_1$ penalization, which is not differentiable, we use the trick borrowed from \citet{andrew2007scalable}: for $a \in \R$, write $|a| = a^+ + a^-$, where $a^+$ and $a^-$ are respectively the positive and negative part of $a$, and add the constraints $a^+ \geq 0$ and $a^- \geq 0$.
Namely, we rewrite the minimization problem~\eqref{eq:minimization-xi_k} as the following differentiable problem with box constraints
\begin{equation}
  \label{eq:sub-problem-xi_k}
  \begin{split}
    \text{minimize}& \quad \quad P_{n, k}^{(w)}(\xi_k^+ - \xi_k^-) + \zeta_{1,k} \big((1 - \eta) \sum_{j=1}^p (\xi_{k,j}^+ + \xi_{k,j}^-) + \dfrac \eta 2 \norm{{\xi_k}^+ - {\xi_k}^-}_2^2 \big) \\
    \text{subject to}& \quad \quad \xi_{k,j}^+ \geq 0 \text{ and } \xi_{k,j}^- \geq 0 \text{ for } j = 1, \ldots, p
  \end{split} 
\end{equation}
where $\xi_k^\pm = (\xi_{k,1}^\pm, \ldots, \xi_{k,p}^\pm)^\top$. The L-BFGS-B solver requires the exact value of the gradient, which is easily given by
\begin{equation}
  \label{eq:grad-xi_k}
  \frac{\partial P_{n,k}^{(w)}(\xi_k)}{\partial \xi_k} = -n^{-1} \sum_{i=1}^n  \big( \hat \pi_{ik}^{\theta^{(w)}} - \pi_{\xi_k}(x_i) \big) x_i.
\end{equation}
In practice, we use the \texttt{Python} solver \texttt{fmin\_l\_bfgs\_b} from \texttt{scipy.optimize}~\citep{virtanen2020scipy}. Similarly to the strategy of increasing $N$ with the iterations $w$ for computational reasons, it is not clever to use a too small solver tolerance in the early steps of the algorithm, where we call tolerance the value for which iterations stop when the stopping criterion is below it. We then decrease the L-BFGS-B tolerance (which we denote $tol$ in Algorithm~\ref{alg:prox-QNEM}) with every increase of $N$.

@@

\begin{equation}
  \label{eq:update_beta}
  {\beta_k}^{(w+1)}= \dfrac{\sum_{i=1}^n\hat \pi_{ik}^{\theta^{(w)}} \Big[U_i^\top I_{n_i} (y_i - V_i\hat \E_{\theta^{(w)}}[ b_i | t_i, \delta_i, y_i])\Big]}{\sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} U_i^\top I_{n_i} U_i}.
\end{equation}
Now, for the update of $\gamma_k^{(w)}$ for $k=0, \ldots, K-1$, we similarly denote
\begin{align*}
  Q^{(w)}_{n,k}(\gamma_k) = -n^{-1} \sum_{i=1}^n &\hat\pi_{ik}^{\theta^{(w)}} \Big[\delta_i \gamma_k^\top \Psi_i -  \exp \big\{\gamma_k^\top \Psi_i\big\} \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \ind{\{t_j^u \leq t_i\}} \Big]
\end{align*}
based on the quantities involved in $\cQ_n(\theta, \theta^{(w)})$ that depend on $\gamma_k$. The integration over the survival process has been replaced with a finite sum over the process evaluated at the distinct event times $(t_1^u, \ldots, t_J^u)$ with $J \in~\N_+$, since $\lambda_0^{(w)}$ defined in~\eqref{eq:update_lambda_0} is always zero except at observed event times. The update for $\gamma_k^{(w)}$ requires to solve the following minimization problem
\begin{equation}
  \label{eq:minimization-gamma_k}
  \gamma_k^{(w+1)} \in \argmin_{\gamma_k \in \R^{L\sA}} Q^{(w)}_{n,k}(\gamma_k) + \zeta_{2,k} \norm{\gamma_k}_{\text{sg} l_1, \tilde{\eta}}.
\end{equation}

@@

\vspace{.5cm}
Lemma~\ref{lemma:prox-sgl1} bellow states that one can easily compute the proximal operator of the sparse group lasso defined in Algorithm~\ref{alg:ISTA}.
\begin{lemma}
\label{lemma:prox-sgl1}
The proximal operator of the sparse group lasso can be expressed as the composition of the group lasso and the lasso proximal operators, that is
\normalfont
\begin{equation*}
  \text{prox}_{\text{sg} l_1,\tilde \eta, \zeta} = \text{prox}_{\zeta \tilde \eta \sum_{l=1}^L \norm{\cdot}_{2,l}} \circ \text{prox}_{\zeta (1 - \tilde \eta) \norm{\cdot}_1}.
\end{equation*}
\end{lemma}
The two proximal operators on the right hand side of Lemma~\ref{lemma:prox-sgl1} are both well known analytically~\citep{bach2011optimization} and tractable. We used the \texttt{Python} library \texttt{copt} for the implementation of proximal gradient descent~\citep{copt} and we propose in our package an implementation for the proximal operator of the sparse group lasso based on Lemma~\ref{lemma:prox-sgl1}.
A proof of Lemma~\ref{lemma:prox-sgl1} is proposed in Appendix~\ref{sec:proof-lemma-prox-sgl1}.
Hence, we update $\gamma_k^{(w)}$ with the following step
\begin{equation}
  \label{eq:gamma-update}
  \gamma_k^{(w+1)} = \text{ISTA}_{\text{sg} l_1}\big(\gamma_k^{(w)}, Q^{(w)}_{n,k}(\cdot), \tilde \eta, \zeta_{2,k}\big)
\end{equation}
Note that the warmstart initialization is left to the user's choice, as well as the ISTA acceleration~\citep{beck2009fast}, since warmstart can diminish acceleration effectiveness~\citep{tibshirani2010proximal}. The gradient of $Q_{n,k}^{(w)}$ is here given by
\begin{align}
  \label{eq:grad-Q}
  &\frac{\partial Q_{n,k}^{(w)}(\gamma_k)}{\partial \gamma_{k,a}^l} = -n^{-1} \sum_{i=1}^n \hat \pi_{ik}^{\theta^{(w)}} \Psi_{i, a}^l  \Big[ \delta_i - \exp \big\{\gamma_k^\top \Psi_i\big\} \sum_{j=1}^J \lambda_0^{(w)}(t_j^u) \ind{\{t_j^u \leq t_i\}} \Big]
\end{align}
for all $a = 1, \ldots, \mathcal{A}$ and $l = 1, \ldots, L$.

For the update of $\lambda_0^{(w)}$, we treat the jump size at each uncensored observed event times as parameters to be estimated~\citep{klein1992semiparametric}, and the closed-form update obtained by cancelling the gradient is given by
\begin{equation}
  \label{eq:update_lambda_0}
  \lambda_0^{(w+1)}(t)= \dfrac{\sum_{i=1}^n \delta_i \ind{\{t=t_i\}}}{\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \exp \big\{{\gamma_k^{(w+1)}}^\top \Psi_i\big\} \ind{\{t_i \geq t\}} },
\end{equation}
for all $t \geq 0$, which is a Breslow like estimator~\citep{breslow1972contribution} adapted to our model.

Finally, for the update of $\phi^{(w)}$ and regarding Table~\ref{table:glm}, we focus on the Gaussian case being particularly useful in practice. It is obtained in closed-form by
\begin{align}
  \label{eq:update_phi}
  \phi_l^{(w+1)} = \big(\sum_{i=1}^n n_i^l \big)^{-1} &\sum_{i=1}^n \sum_{k=0}^{K-1} \hat \pi_{ik}^{\theta^{(w)}} \Big[ \big(y_i^l - U_{il} {\beta_k^l}^{(w+1)}\big)^\top \Big(y_i^l - U_{il} {\beta_k^l}^{(w+1)} \nonumber \\
  & - 2V_{il} \hat \E_{\theta^{(w)}}[ b_i^l | t_i, \delta_i, y_i]\Big) + \Tr\big(V_{il}^\top V_{il}\hat \E_{\theta^{(w)}}[ b_i^l {b_i^l}^\top | t_i, \delta_i, y_i]\big) \Big].
\end{align}
Appendix~\ref{sec:prox-QNEM class} gives some details about the way we implement the equations and their computation in practice.

\paragraph*{Initialization.}
Our algorithm gives a local minimum, and it is clever to choose an initial value $\theta^{(0)}$ close to the final solution $\hat \theta$, so that the number of iterations required to reach convergence is reduced. We then give some details about the starting point $\theta^{(0)}$ of Algorithm~\ref{alg:prox-QNEM}.
For all $k = 0, \ldots, K-1$, we first choose $\xi_k^{(0)} = \mathbf{0}_d$ and ${\gamma_k}^{(0)} = \mathbf{0}_{L\sA}$. Then, we initialize $\lambda_0^{(0)}$ like if there is no subgroup ($\gamma_{0}^{(0)} = \cdots = \gamma_{K-1}^{(0)}$) with a standard Cox proportional hazards regression with time-independent features using the \texttt{Python} library \texttt{tick}. Finally, the longitudinal submodels parameters $\beta_k^{(0)}$, $D^{(0)}$ and $\phi^{(0)}$ are initialized -- again like if there is no subgroup ($\beta_0^{(0)} = \cdots = \beta_{K-1}^{(0)}$) -- using a multivariate linear mixed model with an explicit EM algorithm (details on the algorithm are given in Appendix~\ref{sec:MLMM}, and details on the implementation in Appendix~\ref{sec:MLMM class}), being itself initialized with univariates fits using the \texttt{Python} package \texttt{statsmodels}.

\paragraph*{The prox-QNEM algorithm.}
Algorithm~\ref{alg:prox-QNEM} describes the main steps of the resulting QNEM algorithm to solve the optimization problem~\eqref{eq:optim-pb}.

@@

\begin{algorithm}[!htb]
\DontPrintSemicolon
\caption{prox-QNEM algorithm for the lights model inference}
\label{alg:prox-QNEM}
\KwInput{Training data $\cD_n$; initialize $N=50L$; tuning hyper-parameters $(\zeta_{1,k}, \zeta_{2,k})_{k=0,\ldots,K-1}$}
\KwOutput{Last parameters $\hat \theta \in \R^\vartheta$}
\tcc{Initialization}
Compute the starting parameters $\theta^{(0)} \in \R^\vartheta$ \\
\For{$w = 1,\ldots,$ until convergence}    
{ 
  \tcc{E-step}
  Compute $\hat \Lambda_{ik,\tilde g}^{\theta^{(w)}}$ using~\eqref{eq:Lambda_ik-b_i} for the required functions $\tilde g$\\
  Compute $\hat \pi_{ik}^{\theta^{(w)}}$ using~\eqref{eq:hat_pi_ik-def}\\
  \If{Condition}
  {
      $N \leftarrow N + $\\
      $tol \leftarrow tol - $
  }
  \tcc{Quasi-Newton M-step}
  Update $D^{(w+1)}$ using~\eqref{eq:update_D}\\
  Update $(\xi_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:sub-problem-xi_k} \\
  Update $(\beta_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:update_beta} \\
  Update $(\gamma_k^{(w+1)})_{k=0,\ldots,K-1}$ by solving~\eqref{eq:gamma-update} \\
  Update $\lambda_0^{(w+1)}$ using~\eqref{eq:update_lambda_0}\\
  Update $\phi^{(w+1)}$ using~\eqref{eq:update_phi} \\
}
\textbf{Return:} {$\hat \theta$}
\end{algorithm}
Let us define the relative difference at step $(w+1)$ as
\[\Delta_{\text{rel}}^{(w+1)} = \max_{j=1,\ldots,\vartheta} \Big\{\dfrac{|\theta^{(w+1)}_j - \theta^{(w)}_j|}{|\theta^{(w)}_j| + \epsilon_1 } \Big\}.\]
Then in practice, the convergence for Algorithm~\ref{alg:prox-QNEM} is declared when the condition
\[\Delta_{\text{rel}}^{(w+1)} < \epsilon_2\]
is satisfied for 3 consecutive iterations.

@@

The question now is to choose a discriminative marker rule for risk prediction to be used in the cross-validation procedure for selecting the best regularization hyper-parameters, and of course to be used as final risk prediction after hyper-parameters fine-tuning.
The lights model has the ability to produce prognostic predictions that can be dynamically updated according to the observed trajectory of the features. 
Once the model is trained (so that one obtains $\hat\theta$ from~\eqref{eq:optim-pb} using our prox-QNEM algorithm introduced in Section~\ref{sec:prox-QNEM}), posterior risk classification for subject $i$ and group $k$ can be made through $\hat \pi_{ik}^{\hat\theta}$ (see~\eqref{eq:hat_pi_ik-def}). Similar posterior probabilities have been considered for goodness-of-fit evaluation in other JLCMs models, see~\citet{proust2014joint} for instance.

@@

In this paper, a generalized joint model for high-dimensional multivariate longitudinal data and censored durations (lights) has been introduced, and a new efficient estimation
algorithm (prox-QNEM) has been derived, that considers a penalization of the likelihood in order to perform feature
selection and to prevent overfitting.
